Benefits and risks associated with children’s and adolescents’ interactions with electronic screens: An umbrella review

### 2022-03-07 reports/manuscript_content.Rmd
Second pass of comments

While some forms of screen use (e.g., television viewing) may be detrimental to behaviour and health [@haleScreenTimeSleep2015;@sweetserActivePassiveScreen2012], evidence for other forms of screen exposure (e.g., video games or online communication, such as Zoom) remains less certain and, in some cases, may even be beneficial [@liEarlyChildhoodComputer2004;@warburton2017children]. Thus, according to a Nature Human Behaviour editorial, research to determine the effect of screen exposure on youth is "a defining challenge of the digital age" [@orbenAssociationAdolescentWellbeing2019]. With concerns over the impact of screen use including education, health, social development, and psychological well-being, a broad overview that identifies potential benefits and risks is needed.

Citing the negative effects on health (e.g., increased risk of obesity) and health-related behaviours (e.g., sleep), guidelines from the World Health Organisation [@whoGuidelinesPhysicalActivity2019] and numerous government agencies [@australiangovernmentPhysicalActivityExercise2021;@Canadian24HourMovement2016] and statements by expert groups [@AAPMediaUseSchoolAged2016] have recommended that young people's time spent using electronic media devices for entertainment purposes should be limited. For example, the Australian Government guidelines regarding sedentary behaviour recommend that young children (under the age of two) should not spend any time watching screens. They also recommend that children aged 2-5 years should spend a maximum of one hour engaged in recreational sedentary screen use per day, while children aged 5-12 and adolescents should spend no more than two hours. In contrast, some recent evidence suggests that exposure to electronic entertainment media that exceeds these guidelines (e.g., 3-4 hours per day) may not have meaningful adverse effects on children's behaviour or mental health [@fergusonEverythingModerationModerate2017], and might, in fact, benefit their well-being, as long as this exposure does not reach extreme levels (e.g., 7 hours per day). Some research also indicates that content (e.g., video games vs television programs) may also play an important role in determining the potential benefit or harm of youths' exposure to screen-based media [@przybylskiLargeScaleTestGoldilocks2017]. Indeed, educational screen time is positively related to educational outcomes [@sandersTypeScreenTime2019]. This evidence has led some researchers to argue that a more nuanced approach to screen time guidelines is required [@kayeConceptualMethodologicalMayhem2020].

@@

*Exposure*: We included meta-analyses examining all types of electronic screens including (but not necessarily limited to) television, gaming consoles, computer, tablet, and mobile phones. We also included analyses of all types of content on these devices, including (but not necessarily limited to) recreational content (e.g., television programs, movies, games), homework, and communication (e.g., video chat). In this review we adopted a population-level perspective, meaning that we examined electronic media exposure that occurs during typical daily living activities (e.g., home, school-based electronic media exposure). Consistent with this population-level approach, we excluded technology-based treatments for clinical conditions. However, we included studies examining the effect of screen exposure on non-clinical outcomes (e.g., learning children) for children and youth with a clinical condition. For example, a meta-analysis of the effect of television watching on learning among adolescents diagnosed with depression would be included. However, a meta-analysis of interventions designed to treat clinical depression delivered by a mobile phone app would be excluded.

@@

*Statistical Credibility*. We employed a statistical classification approach to grade the statistical credibility of the effect sizes in the literature. To be considered credible an effect needed to be derived from a combined sample of >1,000 [@papadimitriouUmbrellaReviewEvidence2021] and have non-significant tests of publication bias (i.e., Egger’s test and excess significance test). We performed these analyses, and therefore the review needed to provide usable study-level data in order to be included.

@@

We initially planned to include systematic reviews without meta-analyses in a narrative summary alongside the main meta-analytic findings. However, we determined that combining results from the meta-analyses provided a robust overview. Readers interested in the relevant systematic reviews (i.e., without meta-analysis) can consult the list of references in [Supplementary File 2](https://drive.google.com/file/d/1JhMsySEyfby5MtcUcIVpp0LjDDfTA5IH/view?usp=sharing).

We altered our evidence assessment plan when we identified that, as written, it would treat effects from large reviews, with limited evidence of publication bias, and with consistency of effect in the population as not credible if the effect size was not statistically significant. This would have significantly harmed knowledge gained from our review as it would have restricted our ability to show where the empirical evidence strongly indicated that there was no association between screen time and a given outcome.

@@

freq_exposures <- effects_clean %>%
  group_by(plain_language_exposure) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>%
  mutate(
    plain_language_exposure = rename_effects(plain_language_exposure),
    plain_language_exposure = case_when(
      plain_language_exposure ==
        "lifestyle risk behaviour (at school) intervention" ~
      "screen-based lifestyle risk behaviour interventions (at school)",
      TRUE ~ plain_language_exposure
    ),
    out = paste0(plain_language_exposure, " (n = ", n, ")")
  )

freq_outcomes <- effects_clean %>%
  group_by(plain_language_outcome) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>%
  mutate(
    plain_language_outcome = rename_effects(plain_language_outcome),
    out = paste0(plain_language_outcome, " (n = ", n, ")")
  )

@@

The searches yielded `r format(prisma$data$value[1], big.mark=",")` results, of which `r format(prisma$data$value[2], big.mark=",")` were duplicates. After screening titles and abstracts, we assessed `r format(prisma$data$value[5], big.mark=",")` full-texts for inclusion. `r str_to_sentence(numbers_to_words(prisma$data$value[8], and=TRUE))` met the inclusion criteria and we extracted the data from these meta-analyses. Figure \@ref(fig:prisma) presents the full results of the selection process.

@@

The most frequently reported exposures were `r knitr::combine_words(freq_exposures$out[1:4])`. [Supplementary File 3](https://drive.google.com/file/d/12rB0ihW9Hi1kxM-8AIBfjdCMO6TqP28n/view?usp=sharing) provides a list of all exposures identified. The most frequently reported outcomes were `r knitr::combine_words(freq_outcomes$out[1:5])`. In most cases (`r paste0(sum(freq_combos$n==1), "/", nrow(effects_clean))`), there was only one exposure/outcome combination for an age group, with `r sum(freq_combos$n==2)` appearing twice, and `r sum(freq_combos$n>2)` appearing three or more times. Full characteristics of the included studies are provided in Table \@ref(tab:desctable). Our process yielded `r unique_effects` unique effect/outcome combinations contributed from `r prisma$data$value[9]` reviews. These effects represent the findings of `r format(effects_sum$sum_k, big.mark=",")` primary studies comprised of `r format(effects_sum$sum_n, big.mark=",")` participants.

@@

qual_str <- function(n, totaln, ntype = "", prefix = "(n", suffix = ")") {
  paste0(
    prefix, ntype, "= ", n, "/",
    totaln, ", ", scales::percent(n / totaln, accuracy = 1), suffix
  )
}

@@

The quality of the included meta-analyses was mixed (see Table \@ref(tab:desctable)). Most assessed heterogeneity `r qual_str(n_low_hetero, nrow(qual_data), " low risk ", suffix=" of meta-analyses)")`, reported the characteristics of the included studies `r qual_str(n_low_char, nrow(qual_data), " low risk ")`, and used a comprehensive and systematic search strategy `r qual_str(n_low_search, nrow(qual_data), " low risk ")`. Most reviews did not clearly report if their eligibility criteria were predefined `r qual_str(n_unclear_elig, nrow(qual_data), " unclear ")`. Many papers also did not complete dual independent screening of abstracts and full text `r qual_str(n_high_dualscreen, nrow(qual_data), " high risk ")` or did not clearly report the method of screening `r qual_str(n_unclear_dualscreen, nrow(qual_data), " unclear ")`. A similar trend was observed for dual independent quality assessment `r dualqualstr`. Overall, only `r all_low` meta-analyses were graded as low risk of bias on all criteria. `r all_high_txt` showed high risk of bias on all criteria (if 'unclear' was treated as 'high' then `r all_high_unclear_txt` would be considered all 'high risk').

@@

Among the statistically credible effects, general screen use, television viewing, and video games were all negatively associated with learning. E-books that included narration, as well as touch screen education interventions, and augmented reality education interventions were positively associated with learning. General screen use was negatively associated with literacy outcomes. However, if the screen use involved co-viewing (e.g., watching with a parent), or the content of television programs was educational, the association with literacy was positive (95\% CI, weak evidence). Numeracy outcomes were positively associated with screen-based mathematics interventions and video games that contained numeracy content.

@@

We identified `r n_health_effect` unique outcome-exposure combinations associated with health or health-related behaviour outcomes. We removed `r n_health_noindiv` effects that did not provide individual study-level data, `r n_health_samplesize` effects with samples < 1,000, and `r n_health_faileggers` effects with a significant Egger's test or insufficient studies to conduct the test. No remaining studies showed evidence of excessive significance. Effects not meeting one or more of these standards are presented in [Supplementary File 5](https://drive.google.com/file/d/1-3frrZ3woWjPqkO84Dn3_1CCaE8thIZn/view?usp=sharing). The remaining `r n_health_certain` meta-analytic associations met our criteria for credible evidence and are described below (see also Figure \@ref(fig:healthplot)). These `r n_health_effect` effects came from `r n_health_unique` meta-analytic reviews analysing data from `r health_summary$studies_n` empirical studies with `r format(health_summary$total_n, big.mark=",")` individual participants.

@@

Digital advertising of unhealthy foods—both traditional advertising and video games developed by a brand for promotion—were associated with higher unhealthy food intake. S ocial media use and sexy media exposure were positively associated with risky behaviors (e.g., sexual activity, risk taking, and substance abuse). General screen use was positively associated with depression. Television viewing was negatively correlated with sleep duration, but only at the 95% confidence level (weak evidence). All forms of screen use (general, television, and video games) were positively associated with body composition (e.g., BMI), although the association was smaller for children than for adolescents or for combined populations. Screen-based interventions which target health behaviours appeared effective.

@@

The primary goal of this review was to provide a holistic perspective on the influence of screens on children’s lives across a broad domain of outcomes. We found that when meta-analyses examined general screen use, and did not specify the content, context or device, there was strong evidence showing potentially harmful associations with general learning, literacy, body composition, and depression. However, when meta-analyses included a more nuanced examination of exposures a more complex picture appeared.

@@

Broadly, our findings align with the recommendations of others who suggest that current guidelines may be too simplistic, mischaracterise the strength of the evidence, or do not acknowledge the important nuances of the issue [@elsonPolicyStatementsMedia2019;@ashtonScreenTimeChildren2019;@royalcollegeofpaediatricsandchildhealthHealthImpactsScreen2019]. Our findings suggest that screen use is a complex issue, with associations based not just on duration and device type, but also on the content and the environment in which the exposure occurs. Many current guidelines simplify this complex relationship as something that should be minimised in all instances. We suggest that future guidelines need to embrace the complexity of the issue, to give parents and clinicians specific information to weigh the pros and cons of interactions with screens.

#### Implications for Future Research

Screen use research is big, varied, and rapidly growing. Reviews tended to be general (e.g., all screen time) and even when more targeted (e.g., social media) nuances related to specific content (e.g., Instagram vs Facebook) have not been meta-analysed or have not produced credible evidence.

@@

#### Strengths and Limitations

@@

Our high-level approach limits the feasibility of examining fine-grained details of the individual studies. For example, we did not examine moderators beyond age, nor did we rate the risk of bias for the individual studies. Thus, our assessment of evidence quality was restricted to statistical credibility, rather than a more complete assessment of quality (e.g., GRADE [@guyattGRADEGuidelinesIntroduction2011]). As such, we made decisions regarding the credibility of evidence, where others may have used different thresholds or metrics. For this reason, we provide the complete results in the supplementary material, along with the dataset for others to consider alternative criteria.

#### Conclusions

>>>>

While some forms of screen use (e.g., television viewing) may be detrimental to health and wellbeing [@haleScreenTimeSleep2015;@sweetserActivePassiveScreen2012], evidence for other forms of screen exposure (e.g., video games or online communication, such as Zoom) remains less certain and, in some cases, may even be beneficial [@liEarlyChildhoodComputer2004;@warburton2017children].
Thus, according to a Nature Human Behaviour editorial, research to determine the effect of screen exposure on youth is "a defining challenge of the digital age" [@orbenAssociationAdolescentWellbeing2019].
With concerns over the impact of screen use including education, health, social development, and psychological well-being, a broad overview that identifies potential benefits and risks is needed.

Citing the negative effects of screens on health (e.g., increased risk of obesity) and health-related behaviours (e.g., sleep), guidelines from the World Health Organisation [@whoGuidelinesPhysicalActivity2019] and numerous government agencies [@australiangovernmentPhysicalActivityExercise2021;@Canadian24HourMovement2016] and statements by expert groups [@AAPMediaUseSchoolAged2016] have recommended that young people's time spent using electronic media devices for entertainment purposes should be limited.
For example, the Australian Government guidelines regarding sedentary behaviour recommend that young children (under the age of two) should not spend any time watching screens.
They also recommend that children aged 2-5 years should spend a maximum of one hour engaged in recreational sedentary screen use per day, while children aged 5-12 and adolescents should spend no more than two hours.
In contrast, some recent evidence suggests that exposure to electronic entertainment media that exceeds these guidelines (e.g., 3-4 hours per day) may not have meaningful adverse effects on children's behaviour or mental health, and might, in fact, benefit their well-being, as long as this exposure does not reach extreme levels (e.g., 7 hours per day)[@fergusonEverythingModerationModerate2017].
Some research also indicates that content (e.g., video games vs television programs) plays an important role in determining the potential benefit or harm of youths' exposure to screen-based media [@przybylskiLargeScaleTestGoldilocks2017].
Indeed, educational screen time is positively related to educational outcomes [@sandersTypeScreenTime2019].
This evidence has led some researchers to argue that a more nuanced approach to screen time guidelines is required [@kayeConceptualMethodologicalMayhem2020].

@@

*Exposure*: 
We included meta-analyses examining all types of electronic screens including (but not necessarily limited to) television, gaming consoles, computers, tablets, and mobile phones.
We also included analyses of all types of content on these devices, including (but not necessarily limited to) recreational content (e.g., television programs, movies, games), homework, and communication (e.g., video chat).
In this review we adopted a population-level perspective, meaning that we examined electronic media exposure that occurs during typical daily living activities (e.g., home, school-based electronic media exposure).
Consistent with this population-level approach, we excluded technology-based treatments for clinical conditions.
However, we included studies examining the effect of screen exposure on non-clinical outcomes (e.g., learning) for children and youth with a clinical condition.
For example, a meta-analysis of the effect of television watching on learning among adolescents diagnosed with depression would be included.
However, a meta-analysis of interventions designed to treat clinical depression delivered by a mobile phone app would be excluded.

@@

*Statistical Credibility*.
We employed a statistical classification approach to grade the credibility of the effect sizes in the literature.
To be considered 'credible' an effect needed to be derived from a combined sample of >1,000 [@papadimitriouUmbrellaReviewEvidence2021] and have non-significant tests of publication bias (i.e., Egger’s test and excess significance test).
We performed these analyses, and therefore the review needed to provide usable study-level data in order to be included.

@@

We initially planned to include systematic reviews without meta-analyses in a narrative summary alongside the main meta-analytic findings.
However, we determined that combining results from the meta-analyses allowed readers to compare relative strength of associations more easily.
Readers interested in the relevant systematic reviews (i.e., without meta-analysis) can consult the list of references in [Supplementary File 2](https://drive.google.com/file/d/1JhMsySEyfby5MtcUcIVpp0LjDDfTA5IH/view?usp=sharing).

We altered our evidence assessment plan when we identified that, as written, it could not classify precise evidence of null effects (i.e., from large reviews with low heterogeneity and low risk of publication bias) as 'credible' because a highly-significant *P*-value was a criteria.
This would have significantly harmed knowledge gained from our review as it would have restricted our ability to show where the empirical evidence strongly indicated that there was no association between screen time and a given outcome.

@@

freq_exposures <- effects_clean %>%
  group_by(plain_language_exposure) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>%
  mutate(
    plain_language_exposure = rename_effects(plain_language_exposure),
    plain_language_exposure = case_when(
      plain_language_exposure ==
        "lifestyle risk behaviour (at school) intervention" ~
      "screen-based lifestyle risk behaviour interventions (at school)",
      plain_language_exposure == "general tv programs and movies" ~
      "general TV programs and movies",
      TRUE ~ plain_language_exposure
    ),
    out = paste0(plain_language_exposure, " (*n* = ", n, ")")
  )

freq_outcomes <- effects_clean %>%
  group_by(plain_language_outcome) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>%
  mutate(
    plain_language_outcome = rename_effects(plain_language_outcome),
    plain_language_outcome = case_when(
      plain_language_outcome == "duration sleep" ~ "sleep duration",
      TRUE ~ plain_language_outcome
    ),
    out = paste0(plain_language_outcome, " (*n* = ", n, ")")
  )

@@

The searches yielded `r format(prisma$data$value[1], big.mark=",")` results, of which `r format(prisma$data$value[2], big.mark=",")` were duplicates.
After screening titles and abstracts, we assessed `r format(prisma$data$value[5], big.mark=",")` full-texts for inclusion.
Of those, `r prisma$data$value[8]` met the inclusion criteria and we extracted the data from all of these meta-analyses.
Figure \@ref(fig:prisma) presents the full results of the selection process.

@@

The most frequently reported exposures were `r knitr::combine_words(freq_exposures$out[1:4])`.
[Supplementary File 3](https://drive.google.com/file/d/12rB0ihW9Hi1kxM-8AIBfjdCMO6TqP28n/view?usp=sharing) provides a list of all exposures identified.
The most frequently reported outcomes were `r knitr::combine_words(freq_outcomes$out[1:5])`.
In most cases (`r paste0(sum(freq_combos$n==1), "/", nrow(effects_clean))`), there was only one exposure/outcome combination for an age group, with `r sum(freq_combos$n==2)` appearing twice, and `r sum(freq_combos$n>2)` appearing three or more times.
Full characteristics of the included studies are provided in Table \@ref(tab:desctable).
After removing reviews with duplicate exposure/outcome combinations, our process yielded `r unique_effects` unique effect/outcome combinations contributed from `r prisma$data$value[9]` reviews.
These effects represent the findings of `r format(effects_sum$sum_k, big.mark=",")` primary studies comprised of `r format(effects_sum$sum_n, big.mark=",")` participants.

@@

qual_str <- function(n, totaln, ntype = "", prefix = "(*n*", suffix = ")") {
  paste0(
    prefix, ntype, "= ", n, "/",
    totaln, ", ", scales::percent(n / totaln, accuracy = 1), suffix
  )
}

@@

The quality of the included meta-analyses was mixed (see Table \@ref(tab:desctable)).
Most assessed heterogeneity `r qual_str(n_low_hetero, nrow(qual_data), " low risk ", suffix=" of meta-analyses)")`, reported the characteristics of the included studies `r qual_str(n_low_char, nrow(qual_data), " low risk ")`, and used a comprehensive and systematic search strategy `r qual_str(n_low_search, nrow(qual_data), " low risk ")`.
Most reviews did not clearly report if their eligibility criteria were predefined `r qual_str(n_unclear_elig, nrow(qual_data), " unclear ")`.
Many papers also did not complete dual independent screening of abstracts and full text `r qual_str(n_high_dualscreen, nrow(qual_data), " high risk ")` or did not clearly report the method of screening `r qual_str(n_unclear_dualscreen, nrow(qual_data), " unclear ")`.
A similar trend was observed for dual independent quality assessment `r dualqualstr`.
Overall, only `r all_low` meta-analyses were graded as low risk of bias on all criteria.

@@

Among the statistically credible effects, general screen use, television viewing, and video games were all negatively associated with learning.
E-books that included narration, as well as touch screen education interventions, and augmented reality education interventions were positively associated with learning.
General screen use was negatively associated with literacy outcomes.
However, if the screen use involved co-viewing (e.g., watching with a parent), or the content of television programs was educational, the association with literacy was positive and significant at the 95\% confidence level (weak evidence).
Numeracy outcomes were positively associated with screen-based mathematics interventions and video games that contained numeracy content.

@@

We identified `r n_health_effect` unique outcome-exposure combinations associated with health or health-related behaviour outcomes.
We removed `r n_health_noindiv` effects that did not provide individual study-level data, `r n_health_samplesize` effects with samples < 1,000, and `r n_health_faileggers` effects with a significant Egger's test or insufficient studies to conduct the test.
No remaining studies showed evidence of excessive significance.
Effects not meeting one or more of these standards are presented in [Supplementary File 5](https://drive.google.com/file/d/1-3frrZ3woWjPqkO84Dn3_1CCaE8thIZn/view?usp=sharing).
The remaining `r n_health_certain` meta-analytic associations met our criteria for credible evidence and are described below (see also Figure \@ref(fig:healthplot)).
These `r n_health_certain` effects came from `r n_health_unique` meta-analytic reviews analysing data from `r health_summary$studies_n` empirical studies with `r format(health_summary$total_n, big.mark=",")` individual participants.

@@

Digital advertising of unhealthy foods—both traditional advertising and video games developed by a brand for promotion—were associated with higher unhealthy food intake.
Social media use and sexual content were positively associated with risky behaviors (e.g., sexual activity, risk taking, and substance abuse).
General screen use was positively associated with depression.
Television viewing was negatively correlated with sleep duration, but only at the 95% confidence level (weak evidence).
All forms of screen use (general, television, and video games) were positively associated with body composition (e.g., BMI), although the association was smaller for children than for adolescents or for combined populations.
Screen-based interventions which target health behaviours appeared effective.

@@

The primary goal of this review was to provide a holistic perspective on the influence of screens on children’s lives across a broad domain of outcomes.
We found that when meta-analyses examined general screen use, and did not specify the content, context or device, there was strong evidence showing potentially harmful associations with general learning, literacy, body composition, and depression.
However, when meta-analyses included a more nuanced examination of exposures, a more complex picture appeared.

@@

Broadly, our findings align with the recommendations of others who suggest that current guidelines may be too simplistic, mischaracterise the strength of the evidence, or do not acknowledge the important nuances of the issue [@elsonPolicyStatementsMedia2019;@ashtonScreenTimeChildren2019;@royalcollegeofpaediatricsandchildhealthHealthImpactsScreen2019].
Our findings suggest that screen use is a complex issue, with associations based not just on duration and device type, but also on the content and the environment in which the exposure occurs.
Many current guidelines simplify this complex relationship as something that should be minimised in all instances [@australiangovernmentPhysicalActivityExercise2021;@whoGuidelinesPhysicalActivity2019].
We suggest that future guidelines need to embrace the complexity of the issue, to give parents and clinicians specific information to weigh the pros and cons of interactions with screens.

## Implications for Future Research

Screen use research is big, varied, and rapidly growing.
Reviews tended to be general (e.g., all screen time) and even when more targeted (e.g., social media) nuances related to specific content (e.g., Instagram vs Facebook) have not been meta-analysed or have not produced credible evidence. Fewer than 20\% of the effects identified met our criteria for statistical credibility. Most studies which did not meet our critiera did so because they failed to provide study-level data (or, did not provide sufficent study level data such as including effect sizes but not sample sizes), despite <!--TODO: add reference-->

@@

## Strengths and Limitations

@@

Our high-level approach limits the feasibility of examining fine-grained details of the individual studies.
For example, we did not examine moderators beyond age, nor did we rate the risk of bias for the individual studies.
Thus, our assessment of evidence quality was restricted to statistical credibility, rather than a more complete assessment of quality (e.g., GRADE [@guyattGRADEGuidelinesIntroduction2011]).
As such, we made decisions regarding the credibility of evidence, where others may have used different thresholds or metrics.
For this reason, we provide the complete results in the supplementary material, along with the dataset for others to consider alternative criteria.
In addition, reviews provide only historical evidence which may not keep up with the changing ways children can engage with screens.
While our synthesis of the existing evidence provides good information about how screens might have influenced children in the past, it is difficult to know if these findings will translate to new forms of technology in the future.

## Conclusions
